\Large{Adversarial Variational Optimization of Non-Differentiable Simulators}

### 2017-07-11 tex/paper.tex
Updated experiments section

%Effectively, we are optimizing variational posterior $q(\bftheta | \psi)$ by directly minimizing the 
%divergence of the corresponding posterior predictive distribution $\qxpsi$ and the data distribution $p_r(\bfx)$ without evaluating the unknown joint density $p(\bfx,\bftheta)$.

Effectively, we are optimizing variational posterior $q(\bftheta | \psi)$ by minimizing the 
divergence of the corresponding posterior predictive distribution $\qxpsi$ and the data distribution $p_r(\bfx)$. 
%Effectively, we are performing variational inference through model criticism~\citep{box1980sampling}.
Consequently, the proposed algorithm is a form of variational inference through model criticism~\citep{box1980sampling}, which we denote AVI-MC.

@@

If instead, one is interested in point estimates $\bftheta^*$, then we augment Eqn.~\ref{eqn:vo-ug}
with a regularization term corresponding to the differential entropy $H$ of
the proposal distribution $q(\bftheta | \bfpsi)$. That is,
\begin{equation}
    U_g = \mathbb{E}_{\bftheta \sim q(\bftheta|\bfpsi)} [ {\cal L}_g ] + \gamma H(q(\bftheta|\bfpsi))
\end{equation}
where $\gamma \in \mathbb{R}^+$ is a hyper-parameter controlling the trade-off
between the generator objective and the tightness of the proposal distribution.
For small values of $\gamma$,
proposal distributions with large entropy are not penalized, which results
in learning a smeared variation of the original simulator corresponding to the posterior predictive model. On the other hand,
for large values of $\gamma$, the procedure is encouraged to fit a proposal
distribution with low entropy, which has the effect of concentrating its density
tightly around one or a few $\bftheta$ values. Very large penalties may 
eventually make the optimization unstable, as the variance of $\nabla_\bfpsi \log q(\bftheta_m|\bfpsi)$
typically increases as the entropy of the proposal decreases. \kcnote{maybe, but not obvious why}

@@

The top left plot in Figure~\ref{fig:poisson} illustrates the resulting proposal
distributions $q(\bftheta|\bfpsi)$ after AVO.  For
both $\gamma=0$ and $\gamma=5$, the proposal distributions correctly concentrate
their density around the true parameter value $\log(\lambda^*) = 1.94$. Under
the effect of the positive entropy penalty $H(q(\bftheta|\bfpsi))$,
the proposal distribution for $\gamma=5$ concentrates its mass more tightly,
yielding in this case a more precise inference.  The top right plot compares the
model distributions to the true distribution.  As theoretically expected from
adversarial training, we see that the resulting distributions closely match
the true distribution, with in this case visually slightly better results for the penalized
model.  The bottom plot of Figure~\ref{fig:poisson} shows empirical estimates
of $-U_d$ with respect to the epoch number. For both $\gamma=0$ and $\gamma=5$,
the curves quickly fall towards $0$, which indicates that
$\mathbb{E}_{\tilde{\mathbf{x}} \sim p(\mathbf{x}|\bftheta)}
[d(\tilde{\mathbf{x}};\bfphi)] \approx \mathbb{E}_{\mathbf{x} \sim
p_r(\mathbf{x})} [d(\mathbf{x};\bfphi)]$ and that the critic cannot distinguish
between true and model data. Despite the discreteness and the
non-differentiability of the underlying generator, this confirms that inference
with adversarial variational optimization works.

@@

As a second toy example, we evaluate parameter inference for a generator producing
5-dimensional continuous data, as originally specified in Section 4.2. of
\citep{cranmer2015approximating}. More specifically, we consider the following
generative process:
\begin{itemize}
    \item $\bfz = (z_0, z_1, z_2, z_3, z_4)$, such that
        $z_0 \sim {\cal N}(\mu=\alpha, \sigma=1)$,
        $z_1 \sim {\cal N}(\mu=\beta, \sigma=3)$,
        $z_2 \sim {\text{Mixture}}(\frac{1}{2}\,{\cal N}(\mu=-2, \sigma=1), \frac{1}{2}\,{\cal N}(\mu=2, \sigma=0.5))$,
        $z_3 \sim {\text{Exponential}(\lambda=3)}$, and
        $z_4 \sim {\text{Exponential}(\lambda=0.5)}$;
    \item $\mathbf{x} = R  \bfz$, where $R$ is a fixed
    semi-positive definite $5 \times 5$ matrix defining a fixed projection
    of $\bfz$ into the observed space.
\end{itemize}
We consider observed data generated at the nominal values $\bftheta^* = (\alpha^*=1,\beta^*=-1)$.
The simulator parameters are modeled with a factored Gaussian
proposal distribution $q(\bftheta|\bfpsi) = q(\alpha|\bfpsi) q(\beta|\bfpsi)$, where each component is
initialized with zero mean and unit variance.
Hyper-parameters are set to $M=64$, $n_\text{critic}=100$, $\lambda=0.025$, $\gamma=10$ and
Adam configured with $\alpha=0.01$, $\beta_1=0.9$ and $\beta_2=0.999$.
The network architecture of the critic is the same as in the previous example.

Starting with a proposal distribution $q(\bftheta|\bfpsi)$ largely spread over
the parameter space, as illustrated in the left plot of Figure~\ref{fig:multi},
inference quickly converges towards a proposal distribution whose density
concentrates around the nominal values $\bftheta^*$, as shown in the right plot of Figure~\ref{fig:multi}.
Overall, this example further illustrates and confirms the ability of adversarial
variational optimization for inference with multiple parameters and multidimensional
data, where reliable approximations of $p(\mathbf{x}|\bftheta)$ in a traditional
MLE setting would otherwise be difficult to construct.

@@

As a more challenging example, we now consider a (simplified) simulator from
particle physics for electron--positron collisions resulting in muon--antimuon
pairs ($e^+e^- \rightarrow \mu^+\mu^-$). The simulator approximates the
distribution of observed measurements $\mathbf{x} = \cos(A) \in [-1,1]$, where $A$ is the
polar angle of the outgoing muon with respect  to the originally incoming
electron. This random variable is approximately distributed as
\begin{equation}
    p(\mathbf{x}|E_\text{beam}, G_f) = \frac{1}{Z} \left[ (1 + \mathbf{x}^2) + c(E_\text{beam}, G_f) \mathbf{x} \right]
\end{equation}
where $Z$ is a known normalization constant and $c$ is an asymmetry coefficient
function. Due to the linear term in the expression, the density $p(\mathbf{x} |
E_\text{beam}, G_f)$ exhibits a so-called {\it forward-backward} asymmetry.  Its
size depends on the values of the parameters $E_\text{beam}$ (the beam energy)
and $G_f$ (the Fermi constant) through the coefficient function $c$. Given the
known analytic form of the probability density function, a simulator for
$\mathbf{x}$ can be implemented with rejection sampling~\citep{von195113}. As an
arbitrary number of random drawings may be required before passing the
acceptance threshold, this illustrates how the latent variable
$\bfz$ does not necessarily need to be fixed-sized vector, and how its
density  $p(\bfz|\bftheta)$ may depend on $\bftheta$ in some complicated way
(Section~\ref{sec:problem}).

In this example, we consider observed data generated with $\bftheta^* =
(E_\text{beam}^*=42, G_f^*=0.9)$. The simulator parameters are modeled with a
factored Gaussian proposal distribution $q(\bftheta|\bfpsi) = q(E_\text{beam}|\bfpsi)
q(G_f|\bfpsi)$, where each component is respectively initialized with mean $45$
and $1$ and variance $1$ and $0.01$. Hyper-parameters are set to $M=64$,
$n_\text{critic}=100$, $\lambda=0.0025$ and Adam configured with $\alpha=0.01$,
$\beta_1=0.9$ and $\beta_2=0.999$. As for the first example, we compare
entropy penalties $\gamma=0$ and $\gamma=5$.

The top left plot in Figure~\ref{fig:weinberg} illustrates the resulting
proposal distributions $q(\bftheta|\bfpsi)$ for $\gamma=0$ and $\gamma=5$ after
AVO. We see that the distributions both arrive
in the neighborhood of $\bftheta^*$, with a density  more highly concentrated for
$\gamma=5$ than for $\gamma=0$.  Despite these differences and the relative
distance with $\bftheta^*$, both models closely match the observed data, as shown
in the top right plot of  Figure~\ref{fig:weinberg}, with again slightly better
results for the entropy penalized model. This suggests either a relatively flat
landscape around $\bftheta^*$ or that the observed data can in this case also be
reproduced with a smeared variation $\qxpsi$ of the simulator.
Finally, the bottom plot of Figure~\ref{fig:weinberg} shows that for both
$\gamma=0$ and $\gamma=5$ the variational upper bound $-U_d$ quickly fall
towards $0$, which indicates  convergence towards a distribution that the critic
cannot distinguish from the true distribution. In summary, despite the intricate
relation betwen the observations $\mathbf{x}$, the latent variable $\bfz$
and the parameters $\bftheta$, this example demonstrates the efficacy
of AVO for parameter inference on a black box
simulator.

@@

\section{Related works}

@@

>>>>

Typically in variational inference one would has access to the posterior $p(\bfx | \bftheta)$ (or the joint density $p(\bfx,\bftheta)$), and after optimizing $q(\bftheta | \bfpsi)$ one would criticize the model through the the posterior predictive distribution as advocated by Box~\citep{box1980sampling}.   
Effectively, we are optimizing variational posterior $q(\bftheta | \psi)$ by directly minimizing the 
divergence of the corresponding posterior predictive distribution $\qxpsi$ and the data distribution $p_r(\bfx)$ without evaluating the unknown posterior $p(\bfx | \bftheta)$ or joint density $p(\bfx,\bftheta)$. 
Consequently, the proposed algorithm is a form of variational inference through model criticism~\citep{box1980sampling}, which we denote AVI-MC.

%Effectively, we are optimizing variational posterior $q(\bftheta | \psi)$ by minimizing the 
%divergence of the corresponding posterior predictive distribution $\qxpsi$ and the data distribution $p_r(\bfx)$. 
%%Effectively, we are performing variational inference through model criticism~\citep{box1980sampling}.
%Consequently, the proposed algorithm is a form of variational inference through model criticism~\citep{box1980sampling}, which we denote AVI-MC.

@@

In order to target point estimates $\bftheta^*$, we augment Eqn.~\ref{eqn:vo-ug}
with a regularization term corresponding to the differential entropy $H$ of
the proposal distribution $q(\bftheta | \bfpsi)$. That is,
\begin{equation}
    U_g = \mathbb{E}_{\bftheta \sim q(\bftheta|\bfpsi)} [ {\cal L}_g ] + \gamma H(q(\bftheta|\bfpsi))
\end{equation}
where $\gamma \in \mathbb{R}^+$ is a hyper-parameter controlling the trade-off
between the generator objective and the tightness of the proposal distribution.
For small values of $\gamma$,
proposal distributions with large entropy are not penalized, which results
in learning a smeared variation of the original simulator corresponding to the posterior predictive model. On the other hand,
for large values of $\gamma$, the procedure is encouraged to fit a proposal
distribution with low entropy, which has the effect of concentrating its density
tightly around one or a few $\bftheta$ values.

Very large penalties may 
eventually make the optimization unstable, as the variance of $\nabla_\bfpsi \log q(\bftheta_m|\bfpsi)$
typically increases as the entropy of the proposal decreases. \kcnote{maybe, but not obvious why}

@@

The top left plot in Figure~\ref{fig:poisson} illustrates the resulting proposal
distributions $q(\bftheta|\bfpsi)$ after AVO.  For
both $\gamma=0$ and $\gamma=5$, the proposal distributions correctly concentrate
their density around the true parameter value $\log(\lambda^*) = 1.94$. Under
the effect of the positive entropy penalty $H(q(\bftheta|\bfpsi))$,
the proposal distribution for $\gamma=5$ concentrates its mass more tightly,
yielding in this case a more precise inference.  The top right plot compares the
model distributions to the true distribution.  As theoretically expected from
adversarial training, we see that the resulting distributions closely match
the true distribution, with in this case visually slightly better results for the penalized
model.  The bottom plot of Figure~\ref{fig:poisson} shows empirical estimates
of $-U_d$ with respect to the epoch number. For both $\gamma=0$ and $\gamma=5$,
the curves quickly fall towards $0$, which indicates that
$\mathbb{E}_{\tilde{\mathbf{x}} \sim p(\mathbf{x}|\bftheta)}
[d(\tilde{\mathbf{x}};\bfphi)] \approx \mathbb{E}_{\mathbf{x} \sim
p_r(\mathbf{x})} [d(\mathbf{x};\bfphi)]$ and that the critic cannot distinguish
between true and model data. This confirms that adversarial variational optimization 
works despite the discreteness of the data and lack of access to the 
density $p(\bfx | \bftheta)$ or its gradient.

@@

As a second toy example, we consider a generator producing
5-dimensional continuous data, as originally specified in Section 4.2. of
\citep{cranmer2015approximating}. More specifically, we consider the following
generative process:
\begin{itemize}
    \item $\bfz = (z_0, z_1, z_2, z_3, z_4)$, such that \\
        $z_0 \sim {\cal N}(\mu=\alpha, \sigma=1)$,
        $z_1 \sim {\cal N}(\mu=\beta, \sigma=3)$,
        $z_2 \sim {\text{Mixture}}[\frac{1}{2}\,{\cal N}(\mu=-2, \sigma=1), \\  \frac{1}{2}\,{\cal N}(\mu=2, \sigma=0.5)]$,
        $z_3 \sim {\text{Exponential}(\lambda=3)}$, and
        $z_4 \sim {\text{Exponential}(\lambda=0.5)}$;
    \item $\mathbf{x} = R  \bfz$, where $R$ is a fixed
    semi-positive definite $5 \times 5$ matrix defining a fixed projection
    of $\bfz$ into the observed space.
\end{itemize}
Again, the AVO algorithm does not have access to the density or its gradient, only samples from the generative model
We consider observed data generated at the nominal values $\bftheta^* = (\alpha^*=1,\beta^*=-1)$.
The simulator parameters are modeled with a factored  (mean field) Gaussian
variational distribution $q(\bftheta|\bfpsi) = q(\alpha|\bfpsi) q(\beta|\bfpsi)$, where each component was
initialized with zero mean and unit variance.
Hyper-parameters are set to $M=64$, $n_\text{critic}=100$, $\lambda=0.025$, $\gamma=10$ and
Adam configured with $\alpha=0.01$, $\beta_1=0.9$ and $\beta_2=0.999$.
The network architecture of the critic is the same as in the previous example.

Starting with a proposal distribution $q(\bftheta|\bfpsi)$ largely spread over
the parameter space, as illustrated in the left plot of Figure~\ref{fig:multi},
AVO quickly converges towards a variational distribution whose density
concentrates around the nominal values $\bftheta^*$, as shown in the right plot of Figure~\ref{fig:multi}.
Overall, this example further illustrates and confirms the ability of adversarial
variational optimization for inference with multiple parameters and multidimensional
data, where reliable approximations of $p(\mathbf{x}|\bftheta)$ in a traditional
MLE setting would otherwise be difficult to construct.

@@

As a more realistic example, we now consider a (simplified) simulator from
particle physics for electron--positron collisions resulting in muon--antimuon
pairs ($e^+e^- \rightarrow \mu^+\mu^-$). The simulator approximates the
distribution of observed measurements $\mathbf{x} = \cos(A) \in [-1,1]$, where $A$ is the
polar angle of the outgoing muon with respect  to the originally incoming
electron. Neglecting measurement uncertainty induced from the particle detectors, 
this random variable is approximately distributed as
\begin{equation}\label{eq:simple_weinberg}
    p(\mathbf{x}|E_\text{beam}, G_f) = \frac{1}{Z} \left[ (1 + \mathbf{x}^2) + c(E_\text{beam}, G_f) \mathbf{x} \right]
\end{equation}
where $Z$ is a known normalization constant and $c$ is an asymmetry coefficient
function. Due to the linear term in the expression, the density $p(\mathbf{x} |
E_\text{beam}, G_f)$ exhibits a so-called {\it forward-backward} asymmetry.  Its
size depends on the values of the parameters $E_\text{beam}$ (the beam energy)
and $G_f$ (the Fermi constant) through the coefficient function $c$.

A typical physics simulator for this process includes a more precise treatment of the 
quantum mechanical  $e^+e^- \rightarrow \mu^+\mu^-$ scattering 
using \texttt{MadGraph}~\citep{Alwall:2011uj},  ionization of matter in the 
detector due to the passage of the out-going $\mu^+\mu^-$ particles using 
\texttt{GEANT4}~\citep{Agostinelli:2002hh}, electronic noise and other details of the sensors 
that measure the ionization signal, and the deterministic algorithms that 
estimate the polar angle $A$ based on the sensor readouts. The simulation of 
this process is highly non-trivial as is the space of latent variables $\mathcal{Z}$.

%Given the
%known analytic form of the probability density function, a simulator for
%$\mathbf{x}$ can be implemented with rejection sampling~\citep{von195113}. As an
%arbitrary number of random drawings may be required before passing the
%acceptance threshold, this illustrates how the latent variable
%$\bfz$ does not necessarily need to be fixed-sized vector, and how its
%density  $p(\bfz|\bftheta)$ may depend on $\bftheta$ in some complicated way
%(Section~\ref{sec:problem}).

In this example, we consider observed data generated with the simplified generator of Eq.~\ref{eq:simple_weinberg}
using $\bftheta^* = (E_\text{beam}^*=42, G_f^*=0.9)$. The simulator parameters are modeled with a
factored (mean field) Gaussian variational distribution $q(\bftheta|\bfpsi) = q(E_\text{beam}|\bfpsi)
q(G_f|\bfpsi)$, where each component is respectively initialized with mean $45$
and $1$ and variance $1$ and $0.01$. Hyper-parameters are set to $M=64$,
$n_\text{critic}=100$, $\lambda=0.0025$ and Adam configured with $\alpha=0.01$,
$\beta_1=0.9$ and $\beta_2=0.999$. As with the first example, we compare
entropy penalties $\gamma=0$ and $\gamma=5$.

The top left plot in Figure~\ref{fig:weinberg} illustrates the resulting
proposal distributions $q(\bftheta|\bfpsi)$ for $\gamma=0$ and $\gamma=5$ after
AVO. We see that the distributions both arrive
in the neighborhood of $\bftheta^*$, with a density  more highly concentrated for
$\gamma=5$ than for $\gamma=0$.  Despite these differences and the relative
distance with $\bftheta^*$, both models closely match the observed data, as shown
in the top right plot of  Figure~\ref{fig:weinberg}, with again slightly better
results for the entropy penalized model. This suggests either a relatively flat
landscape around $\bftheta^*$ or that the observed data can in this case also be
reproduced with the posterior predictive distribution $\qxpsi$.
Finally, the bottom plot of Figure~\ref{fig:weinberg} shows that for both
$\gamma=0$ and $\gamma=5$ the variational upper bound $-U_d$ quickly fall
towards $0$, which indicates  convergence towards a distribution that the critic
cannot distinguish from the true distribution. 
%In summary, despite the intricate
%relation between the observations $\mathbf{x}$, the latent variable $\bfz$
%and the parameters $\bftheta$, this example demonstrates the efficacy
%of AVO for parameter inference on a black box
%simulator.

@@

\section{Related work}

@@

\kcnote{proposal distribution $\to$ variational distribution}
